---
title: "RR CP 2"
author: "Yigit Ozan Berk"
date: "5/20/2019"
output: html_document
---

Synopsis : This report tries to answer some basic questions about severe weather events in US between 1950 and 2011. 

1. Across the United States, which types of events (EVTYPE) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

The report consists of two parts, *Data Processing* and *Results*. The first part shows how the data is organized for analysis. The second part shows how the final results are obtained through processed data.

-----------------------------

Contents : 
1. Data Processing
2. Results


# Data Processing

Initiation 
```{r}
library(ggplot2)
library(dplyr)
```

Read the data file

```{r}
wdata <- read.csv("repdata-data-StormData.csv.bz2")
```

-Fatality codes : REMARKS column. need to modify string.
-Fatality N : FATALITIES column
-Beginning Date : BGN_DATE %m/%d/%Y 0:00:00
-Beginnign Time : BGN_TIME %
-End Date : END_DATE
-End Time : END_TIME
-Time Zone : TIME_ZONE - 3 letter string
-State : STATE__, STATE
-Event Type : EVTYPE

# Results  **Not more than 3 figures(each figure can have multiple plots)**

